Randomized trace estimation for parameter-dependent matrices applied to spectral density approximation

### 2025-02-21 paper/figures/numerical-rank.tex
generate plots which illustrate improvements
rework numerical results section and numerical rank explanation

\begin{tikzpicture}
    \fill[lightblue] (-2, -0.15) rectangle (1.5, 0.15);
    \draw[thick, ->] (-5, 0) to (5, 0);
    \fill[darkblue] (-4, 0) circle (0.075) node[above] {$\lambda_{(7)}$};
    \fill[darkblue] (-3, 0) circle (0.075) node[above] {$\lambda_{(5)}$};
    \fill[darkblue] (-2.3, 0) circle (0.075) node[above] {$\lambda_{(4)}$};
    \fill[darkblue] (-1, 0) circle (0.075) node[above] {$\lambda_{(2)}$};
    \fill[darkblue] (0.3, 0) circle (0.075) node[above] {$\lambda_{(1)}$};
    \fill[darkblue] (1, 0) circle (0.075) node[above] {$\lambda_{(3)}$};
    \fill[darkblue] (2.8, 0) circle (0.075) node[above] {$\lambda_{(6)}$};
    \fill[darkblue] (3.6, 0) circle (0.075) node[above] {$\lambda_{(8)}$};
    \fill[darkblue] (4.4, 0) circle (0.075) node[above] {$\lambda_{(9)}$};
    \draw[darkblue, ultra thick] (-0.155, 0.15) to (-0.155, -0.15) node[below] {$t$};
    \draw[darkblue, thick] (-2, 0.15) to (-2, -0.15) node[below] {$t - d_{\varepsilon, \sigma}$};
    \draw[darkblue, thick] (1.5, 0.15) to (1.5, -0.15) node[below] {$t + d_{\varepsilon, \sigma}$};
\end{tikzpicture}

>>>>

\begin{tikzpicture}
    \fill[black!15!white] (-2.0, -0.15) rectangle (2.0, 0.15);
    \draw[thick, ->] (-5, 0) to (5, 0);
    \fill[black] (-4, 0) circle (0.075) node[below] {$\lambda_{(7)}$};
    \fill[black] (-3, 0) circle (0.075) node[below] {$\lambda_{(5)}$};
    \fill[black] (-2.35, 0) circle (0.075) node[below] {$\lambda_{(4)}$};
    \fill[black] (-0.8, 0) circle (0.075) node[below] {$\lambda_{(2)}$};
    \fill[black] (0.3, 0) circle (0.075) node[below] {$\lambda_{(1)}$};
    \fill[black] (1, 0) circle (0.075) node[below] {$\lambda_{(3)}$};
    \fill[black] (3, 0) circle (0.075) node[below] {$\lambda_{(6)}$};
    \fill[black] (4.2, 0) circle (0.075) node[below] {$\lambda_{(8)}$};
    \fill[black] (4.8, 0) circle (0.075) node[below] {$\lambda_{(9)}$};
    \fill[black] (-4, 0) circle (0.075) node[above] {$\sigma_{7}$};
    \fill[black] (-3, 0) circle (0.075) node[above] {$\sigma_{5}$};
    \fill[black] (-2.35, 0) circle (0.075) node[above] {$\sigma_{4}$};
    \fill[black] (-0.8, 1.3) circle (0.075) node[above left] {$\sigma_{2}$};
    \fill[black] (0.3, 2.28) circle (0.075) node[above right] {$\sigma_{1}$};
    \fill[black] (1, 0.91) circle (0.075) node[above right] {$\sigma_{3}$};
    \fill[black] (3, 0) circle (0.075) node[above] {$\sigma_{6}$};
    \fill[black] (4.2, 0) circle (0.075) node[above] {$\sigma_{8}$};
    \fill[black] (4.8, 0) circle (0.075) node[above] {$\sigma_{9}$};
    \draw[black, ultra thick] (-0.05, 0.15) to (-0.05, -0.55) node[below] {$t$};
    \draw[black, thick] (-2.0, 0.15) to (-2.0, -0.55) node[below] {$t - d$};
    \draw[black, thick] (2.0, 0.15) to (2.0, -0.55) node[below] {$t + d$};
    \draw[black, thick, dashed] (-0.8, 0) to (-0.8, 1.3);
    \draw[black, thick, dashed] (0.3, 0) to (0.3, 2.28);
    \draw[black, thick, dashed] (1, 0) to (1, 0.91);
    \draw[black!60!white, thick] (-5.000, 0.000) to (-4.899, 0.000) to (-4.798, 0.000) to (-4.697, 0.000) to (-4.596, 0.000) to (-4.495, 0.000) to (-4.394, 0.000) to (-4.293, 0.000) to (-4.192, 0.000) to (-4.091, 0.000) to (-3.990, 0.000) to (-3.889, 0.000) to (-3.788, 0.000) to (-3.687, 0.000) to (-3.586, 0.000) to (-3.485, 0.000) to (-3.384, 0.000) to (-3.283, 0.000) to (-3.182, 0.000) to (-3.081, 0.000) to (-2.980, 0.000) to (-2.879, 0.001) to (-2.778, 0.001) to (-2.677, 0.002) to (-2.576, 0.003) to (-2.475, 0.005) to (-2.374, 0.009) to (-2.273, 0.014) to (-2.172, 0.022) to (-2.071, 0.034) to (-1.970, 0.052) to (-1.869, 0.076) to (-1.768, 0.110) to (-1.667, 0.155) to (-1.566, 0.215) to (-1.465, 0.293) to (-1.364, 0.389) to (-1.263, 0.508) to (-1.162, 0.649) to (-1.061, 0.812) to (-0.960, 0.995) to (-0.859, 1.196) to (-0.758, 1.408) to (-0.657, 1.625) to (-0.556, 1.836) to (-0.455, 2.033) to (-0.354, 2.206) to (-0.253, 2.346) to (-0.152, 2.443) to (-0.051, 2.494) to (0.051, 2.494) to (0.152, 2.443) to (0.253, 2.346) to (0.354, 2.206) to (0.455, 2.033) to (0.556, 1.836) to (0.657, 1.625) to (0.758, 1.408) to (0.859, 1.196) to (0.960, 0.995) to (1.061, 0.812) to (1.162, 0.649) to (1.263, 0.508) to (1.364, 0.389) to (1.465, 0.293) to (1.566, 0.215) to (1.667, 0.155) to (1.768, 0.110) to (1.869, 0.076) to (1.970, 0.052) to (2.071, 0.034) to (2.172, 0.022) to (2.273, 0.014) to (2.374, 0.009) to (2.475, 0.005) to (2.576, 0.003) to (2.677, 0.002) to (2.778, 0.001) to (2.879, 0.001) to (2.980, 0.000) to (3.081, 0.000) to (3.182, 0.000) to (3.283, 0.000) to (3.384, 0.000) to (3.485, 0.000) to (3.586, 0.000) to (3.687, 0.000) to (3.788, 0.000) to (3.889, 0.000) to (3.990, 0.000) to (4.091, 0.000) to (4.192, 0.000) to (4.293, 0.000) to (4.394, 0.000) to (4.495, 0.000) to (4.596, 0.000) to (4.697, 0.000) to (4.798, 0.000) to (4.899, 0.000) to (5.000, 0.000);
    \node[black!60!white] at (-1.3, 2.3) {$g_{\sigma}(t - \cdot)$};
\end{tikzpicture}

### 2025-02-21 paper/intro.tex
rework numerical results section and numerical rank explanation

\paragraph{New contributions.} We analyze three well-established randomized trace estimators when they are applied to parameter-dependent matrices: The Girard-Hutchinson estimator \cite{girard-1989-fast-montecarlo, hutchinson-1990-stochastic-estimator}, the trace of the Nyström low-rank approximation \cite{gittens-2013-revisiting-nystrom}, and the Nyström++ estimator \cite{persson-2022-improved-variants}. Interestingly, we can reuse the same randomization for computing the estimate at each value of the parameter $t$, which makes the estimators scale favorably with the number of parameter evaluations. We propose certain modifications to the methods from \cite{lin-2017-randomized-estimation} for approximating the smoothed spectral density $\phi_{\sigma}$ \refequ{equ:smoothed-spectral-density}. In particular, we use a more rigorous approach for expanding matrix functions in terms of Chebyshev polynomials and unify all of their methods in a single algorithm, with the desirable side effect of speeding up their \enquote{spectrum sweeping} method by orders of magnitude. Based on the analysis of the randomized trace estimators for parameter-dependent matrices, we provide bounds on the approximation error for our method.

>>>>

\paragraph{New contributions.} We analyze three well-established randomized trace estimators when they are applied to parameter-dependent matrices: The Girard-Hutchinson estimator \cite{girard-1989-fast-montecarlo, hutchinson-1990-stochastic-estimator}, the trace of the Nyström low-rank approximation \cite{gittens-2013-revisiting-nystrom}, and the Nyström++ estimator \cite{persson-2022-improved-variants}. Interestingly, we can reuse the same randomization for computing the estimate at each value of the parameter $t$, which makes the estimators scale favorably with the number of parameter evaluations. We propose certain modifications to the methods from \cite{lin-2017-randomized-estimation} for approximating the smoothed spectral density $\phi_{\sigma}$ \refequ{equ:smoothed-spectral-density}. In particular, we use a more rigorous approach for approximating matrix functions in terms of Chebyshev polynomials and unify all of the methods of \cite{lin-2017-randomized-estimation} in a single algorithm, with the desirable side effect of speeding up the \enquote{spectrum sweeping} method by orders of magnitude. Based on the analysis of the randomized trace estimators for parameter-dependent matrices, we provide bounds on the approximation error for our method.

### 2025-02-21 paper/numresults.tex
rework numerical results section and numerical rank explanation

In this last section, we test our algorithm in multiple scenarios. First, we numerically confirm our theoretical convergence bounds by approximating the spectral density of a Hamiltonian arising from electronic structure interaction. Subsequently, we compare our method to the Krylov-aware stochastic trace estimator, once on the electronic structure example, and once for estimating the trace of a more general parameter-dependent matrix appearing in statistical thermodynamics. In the end, we demonstrate that our algorithm can also be used to monitor the optimization process of a neural network.

All our implementations are written in Python 3.12.3 using the packages NumPy 2.0.0 and SciPy 1.14.1. They are executed on a single thread of a GitHub-hosted Ubuntu runner with a 64-bit processor and 16 GB of RAM. \textcolor{red}{??? Provide all parameter choices algorithmic choices common toa all experiments here. For example, the dismissal of non-negative approx ???}
(evtl. unless otherwise mentioned 
$n_t$ sampling points uniformly distributed.
The error is calculcated with the composite midpoint rule using this sampling points.
For transforming the interval to $[-1,1]$ exterior eigenvalues of $A$ are estimated using the Lanczos method (Python's ....).

@@

Our first numerical example comes from electronic structure interaction \textcolor{red}{??? Provide more specific reference ???} \cite{lin-2017-randomized-estimation}, more precisely from the second order finite difference discretization of the Hamiltonian
\begin{equation}
    \mathcal{H} = - \Delta + V
    \label{equ:5-experiments-electronic-hamiltonian}
\end{equation}
in three dimensions. The potential $V$ interacting with the electrons is generated by Gaussian wells
\begin{equation}
    v(r) = v_0 e^{-\lambda r^2}
    \label{equ:5-experiments-gaussian-cell}
\end{equation}
with $v_0 = -4$ and $\lambda = 8$, centered in cells of side length $L=6$ which are stacked $n_c \in \mathbb{N}$ times in each spatial dimension (cf. \reffig{fig:gaussian-well}). The grid width is fixed to be $h=0.6$. For example, for $n_c = 1$, this leads to a matrix of size \textcolor{red}{??? ???}. This is an idealized model for the interaction of nuclei on a regular grid with electrons for a $k$-vector in the center of the first Brillouin zone. The distribution of the eigenvalues of the Hamiltonian --- its spectral density --- allows us to interpret the system's energy levels.

@@

While theoretically an important tool, for the remainder of this section, we waive using the non-negative Chebyshev approximation, since the (slight) indefiniteness of $g_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})$ \refequ{equ:matrix-approximation} does not seem to matter when computing the Nyström++ estimator in finite precision arithmetic. We approximate the integrals for computing the $L^1$-errors in \refsec{sec:analysis} using composite midpoint quadrature with $n_t \in \mathbb{N}$ uniformly distributed nodes. For a fixed total number of random vectors $n_{\mtx{\Psi}} + n_{\mtx{\Omega}}$, we analyze the convergence behavior of the Chebyshev-Nyström++ method when run on the $1000 \times 1000$ matrix resulting from the finite difference discretization of the Hamiltonian for $n_c = 1$ described above (see \reffig{fig:convergence}).

\begin{figure}[!htpb]
    \centering
    \input{plots/convergence.pgf}
    \caption{For increasing values of $n_{\mtx{\Psi}} + n_{\mtx{\Omega}}$ but fixed $m$, we apply the Chebyshev-Nyström++ method to the Hamiltonian matrix described in \refsec{subsec:hamiltonian}. We plot the $L^1$-approximation error for $\sigma=0.005$ on the model problem with $n_c = 1$. The $L^1$-error is approximated by a quadrature with $n_t = 100$ evenly spaced nodes in $[-1, 1]$.}
    \label{fig:convergence}
\end{figure}

Indeed, we observe that for $n_{\mtx{\Omega}} = 0$, we require $n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-2})$ to achieve an error of order $\varepsilon$, as shown in \refthm{thm:hutchinson}. Because the eigenvalues of the matrix are quite uniformly distributed, we observe the stronger convergence of the Nyström approximation discussed in the end of \refsec{sec:application} once $n_{\mtx{\Omega}}$ exceeds the numerical rank \refequ{equ:gaussian-kernel-numerical-rank-uniform}, which in this case would be somewhat larger than $30$.

@@

\begin{figure}[!htpb]
    \centering
    \input{plots/distribution.pgf}
    \caption{The Chebyshev-Nyström++ method is applied to the model problem with $n_c = 1$ for different ways of allocations a total of $n_{\mtx{\Psi}} + n_{\mtx{\Omega}}=80$ random vectors to either the Nystr\"om low-rank approximation or the Girard-Hutchinson trace estimation for the Gaussian smoothing kernel with multiple different values of the smoothing parameter. We make the approximation error made in the Chebyshev approximation negligible by rescaling $m=16 / \sigma$ (based on \reflem{lem:non-negative-chebyshev-error}). The $L^1$-error is approximated by a quadrature with $n_t = 100$ evenly spaced nodes in $[-1, 1]$.}
    \label{fig:distribution}
\end{figure}

@@

The Krylov-aware stochastic trace estimation \cite{chen-2023-krylovaware-stochastic} is also well suited for approximating the trace of a matrix function $f(\mtx{A}, t)$ which separately depends on a parameter $t$. It also samples two Gaussian random matrices $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ and $\mtx{\Psi} \in \mathbb{R}^{n \times n_{\mtx{\Psi}}}$. $\mtx{\Omega}$ is used to extract a low-rank approximation to $f(\mtx{A}, t)$ from a block-Krylov subspace of $\mtx{A}$ generated from $r$ iterations of the block-Lanczos method with reorthogonalization and subsequently $q$ without. $\mtx{\Psi}$ is then used to estimate the trace of the approximation residual with $r$ iterations of the Lanczos method.

@@

To demonstrate that the Chebyshev-Nyström++ estimator is also effective on problems unrelated to spectral density estimation, we consider the problem of approximating $\Trace(\exp(-\beta \mtx{A}))$ for the Hamiltonian $\mtx{A}$ of a planar Heisenberg spin chain \cite{chen-2023-krylovaware-stochastic}. We compute the pointwise error from the analytic solution as a function of the temperature $\beta^{-1}$ and compare it to the error achieved by the Krylov-aware trace estimator in \reffig{fig:krylov-aware-spin}.

Unlike in the spectral density example from \refsec{subsec:hamiltonian}, the numerical rank of the matrix we approximate wildly changes with the parameter $\beta$. Since the Chebyshev-Nyström++ method requires a fixed allocation of random vectors to the low-rank approximation for all values of the parameter $\beta$. Hence, both for large $\beta$, where the matrix $\exp(-\beta \mtx{A})$ is low-rank, and for small $\beta$, where the singular values only decay slowly, we cannot expect a good approximation.

@@

To demonstrate that our algorithm can effectively be applied in the setting of neural network optimization, we approximate the spectral density of a small fully connected convolutional neural network with 6782 parameters. We train this network on the handwritten digit classification task given by the MNIST dataset\footnote{Handwritten digit classification; taken from \url{http://yann.lecun.com/exdb/mnist}.} in PyTorch 2.4.1 in a standard way. We plot the approximation of the spectral density of the Hessian matrix of the untrained neural network, and at different stages of training in \reffig{fig:hessian-density}, as well as the corresponding mean squared error loss. It can well be observed that the eigenvalues creep towards zero as the training proceeds. Furthermore, despite the loss steadily decreasing, the presence of large eigenvalues in some epochs indicates a sharp minimum of the network, hence, unfavorable generalization properties.

\begin{figure}
    \begin{minipage}[c]{.49\linewidth}
        \centering
        \input{plots/hessian_density.pgf}
    \end{minipage}\hfill%
    \begin{minipage}[c]{.49\linewidth}
        \centering
        \input{plots/hessian_density_loss.pgf}
    \end{minipage}
    \caption{The mean squared error training loss (right) and the corresponding approximate spectral density of the Hessian matrix of a fully connected convolutional neural network in different epochs of training on the MNIST dataset (left). Computed with the Chebyshev-Nyström++ method (\refalg{alg:nystrom-chebyshev-pp}) on $n_{\mtx{\Omega}} = 10$, $n_{\mtx{\Psi}} = 10$, $m = 1000$, and $\sigma = 0.005$.}
    \label{fig:hessian-density}
\end{figure}

>>>>

In this last section, we test the Chebyshev-Nyström++ algorithm in multiple scenarios leading from electronic structure interaction, statistical thermodynamics, and neural network optimization.

While theoretically an important tool, for the remainder of this section, we waive using the non-negative Chebyshev approximation, since the (slight) indefiniteness of $g_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})$ \refequ{equ:matrix-approximation} does not seem to matter when computing the Nyström++ estimator in finite precision arithmetic. Unless otherwise mentioned, we transform the matrices such that all their eigenvalues are contained in $[-1, 1]$ by approximating the spectrum with NumPy's Hessian eigenvalue solver. We compute the spectral densities at $n_t = 100$ uniformly distributed values of the parameter $t$ in $[-1, 1]$. The integral involved in computing the $L^1$-errors is approximated with the composite midpoint quadrature rule using the $n_t$ values of $t$ as nodes. Both parameters introduced at the end \refsec{subsubsec:chebyshev-nystrom-implementation},the eigenvalue truncation threshold and the parameter for detecting a vanishing spectral density, are both set to $10^{-5}$.

All our implementations are written in Python 3.12.3 using the packages NumPy 2.0.0 and SciPy 1.14.1. They are executed on a single thread of a GitHub-hosted Ubuntu runner with a 64-bit processor and 16 GB of RAM.

@@

Our first numerical example comes from electronic structure interaction \cite[Section 6]{lin-2017-randomized-estimation}, more precisely from the second order finite difference discretization of the Hamiltonian
\begin{equation}
    \mathcal{H} = - \Delta + V
    \label{equ:5-experiments-electronic-hamiltonian}
\end{equation}
in three dimensions. The potential $V$ interacting with the electrons is generated by Gaussian wells
\begin{equation}
    v(r) = v_0 e^{-\lambda r^2}
    \label{equ:5-experiments-gaussian-cell}
\end{equation}
with $v_0 = -4$ and $\lambda = 8$, centered in cells of side length $L=6$ which are stacked $n_c \in \mathbb{N}$ times in each spatial dimension (cf. \reffig{fig:gaussian-well}). The grid width is fixed to be $h=0.6$. For example, for $n_c = 1$, this leads to a matrix of size $1000 \times 1000$. This is an idealized model for the interaction of nuclei on a regular grid with electrons for a $k$-vector in the center of the first Brillouin zone. The distribution of the eigenvalues of the Hamiltonian --- its spectral density --- allows us to interpret the system's energy levels.

@@

For a fixed total number of random vectors $n_{\mtx{\Psi}} + n_{\mtx{\Omega}}$, we analyze the convergence behavior of the Chebyshev-Nyström++ method when run on the $1000 \times 1000$ matrix resulting from the finite difference discretization of the Hamiltonian for $n_c = 1$ described above (see \reffig{fig:convergence}). As a reference, we use the eigenvalues computed by NumPy's Hermitian eigenvalue solver applied to the matrix $\mtx{A}$.

\begin{figure}[ht]
    \centering
    \input{plots/convergence.pgf}
    \caption{For increasing values of $n_{\mtx{\Psi}} + n_{\mtx{\Omega}}$ but fixed $m$, we apply the Chebyshev-Nyström++ method to the Hamiltonian matrix described in \refsec{subsec:hamiltonian}. We plot the $L^1$-approximation error for $\sigma=0.005$ on the model problem with $n_c = 1$.}
    \label{fig:convergence}
\end{figure}

Indeed, we observe that for $n_{\mtx{\Omega}} = 0$, we require $n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-2})$ to achieve an error of order $\varepsilon$, as shown in \refthm{thm:hutchinson}. Because the eigenvalues of the matrix are quite uniformly distributed, we observe the stronger convergence of the Nyström approximation discussed in the end of \refsec{sec:application} once $n_{\mtx{\Omega}}$ is large enough.

@@

\begin{figure}[ht]
    \centering
    \input{plots/distribution.pgf}
    \caption{For a range of values of $\sigma$, the Chebyshev-Nyström++ method is applied to the Hamiltonian matrix described in \refsec{subsec:hamiltonian} with $n_c = 1$ for different ways of allocating a total of $n_{\mtx{\Psi}} + n_{\mtx{\Omega}}=80$ random vectors to either the Nystr\"om low-rank approximation or the Girard-Hutchinson trace estimation. We make the approximation error committed by the Chebyshev approximation negligible by rescaling $m=16 / \sigma$ (based on \reflem{lem:non-negative-chebyshev-error}).}
    \label{fig:distribution}
\end{figure}

@@

The Krylov-aware stochastic trace estimator \cite[Algorithm 3.1]{chen-2023-krylovaware-stochastic} is also suitable in the context of spectral density estimation. It also samples two Gaussian random matrices $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ and $\mtx{\Psi} \in \mathbb{R}^{n \times n_{\mtx{\Psi}}}$. $\mtx{\Omega}$ is used to extract a low-rank approximation to $g_{\sigma}(t \mtx{I}_n - \mtx{A})$ by running the block-Lanczos algorithm on $\mtx{A}$ with starting block $\mtx{\Omega}$ for $r$ iterations with reorthogonalization and subsequently $q$ without. The columns of $\mtx{\Psi}$ are then used to estimate the trace of the approximation residual with $r$ iterations of the Lanczos algorithm on $\mtx{A}$.

@@

@@

To demonstrate that our algorithm can effectively be applied in the setting of neural network optimization, we approximate the spectral density of a small fully connected convolutional neural network with 6782 parameters. We train this network on the handwritten digit classification task given by the MNIST dataset\footnote{Handwritten digit classification; taken from \url{http://yann.lecun.com/exdb/mnist}.} in PyTorch 2.4.1 in a standard way. We use the bound from \cite[Theorem 1]{zhou-2011-bounding-spectrum} to estimate a bound on the spectrum.

We plot the approximation of the spectral density of the Hessian matrix of the untrained neural network, and at different stages of training in \reffig{fig:hessian-density}, as well as the corresponding mean squared error loss. It can well be observed that the eigenvalues creep towards zero as the training proceeds. Furthermore, despite the loss steadily decreasing, the presence of relatively large eigenvalues in some epochs indicates a sharp minimum of the network, hence, unfavorable generalization properties.

\begin{figure}
    \begin{minipage}[c]{.49\linewidth}
        \centering
        \input{plots/hessian_density.pgf}
    \end{minipage}\hfill%
    \begin{minipage}[c]{.49\linewidth}
        \centering
        \input{plots/hessian_density_loss.pgf}
    \end{minipage}
    \caption{The mean squared error training loss (right) and the corresponding approximate spectral density of the Hessian matrix of a fully connected convolutional neural network in different epochs of training on the MNIST dataset (left). The spectral density is approximated in $n_t=150$ uniformly spaced points using the Chebyshev-Nyström++ method (\refalg{alg:nystrom-chebyshev-pp}) with parameters $n_{\mtx{\Omega}} = 10$, $n_{\mtx{\Psi}} = 10$, $m = 1000$, and $\sigma = 0.005$.}
    \label{fig:hessian-density}
\end{figure}

\subsection{Approximation of the partition function of a quantum system}
\label{subsec:spin-chain}

To demonstrate that the Chebyshev-Nyström++ estimator is also effective on problems unrelated to spectral density estimation, we consider the problem of approximating the partition function $\Trace(\exp(-\beta \mtx{A}))$ for the Hamiltonian $\mtx{A}$ of a planar Heisenberg spin chain \cite[Section 5.1]{chen-2023-krylovaware-stochastic}. We compute the pointwise error from the analytic solution as a function of the temperature $\beta^{-1}$ and compare it to the error achieved by the Krylov-aware trace estimator in \reffig{fig:krylov-aware-spin}.

Unlike in the spectral density example from \refsec{subsec:hamiltonian}, where the numerical rank of the matrix stays low for all $t$, it changes wildly with the parameter $\beta$. The Chebyshev-Nyström++ method requires a fixed allocation of random vectors to the low-rank approximation for all values of the parameter $\beta$. Hence, both for large $\beta$, where the matrix $\exp(-\beta \mtx{A})$ is low-rank, and for small $\beta$, where the singular values only decay slowly, we cannot expect a good approximation.
